---
title: "Creating new variables"
author: "Hector Antonio Vazquez Brust"
date: "October 9, 2015"
output: html_document
---

##Boston neighborhoods: A more nuanced view on assessed value


This time we'll create some new variables to fix problems with our dataset, and also to facilitate a deeper analysis.


```{r, echo=FALSE}
options(scipen = 99) # this prevents R from displaying large numbers with scientific notation

library(ggplot2)
library(plyr)

TAdata <- read.csv('data/Tax Assessor 2015 - Data.csv') #load the Tax Assesor dataset

```



As we've seen before, our dataset holds many properties whose year of construction is listed as zero:

```{r, warning=FALSE}
ggplot(TAdata, aes(x = YR_BUILT)) + geom_histogram(binwidth = 10) + ggtitle("Boston: building construction dates") + labs(x="Year",y="Count")

```


We will fix that by creating a "revised" variable, which will hold a "NA" where the YR_BUILT column lists a "0", and otherwise it will reflect the construction year: 

```{r}
TAdata <- transform(TAdata, YR_BUILT_REVISED = ifelse(YR_BUILT == 0, NA, YR_BUILT))
```

Let's check how it looks now:

```{r, warning=FALSE}
ggplot(TAdata, aes(x = YR_BUILT_REVISED)) + geom_histogram(binwidth = 10) + ggtitle("Boston: building construction dates") + labs(x="Year",y="Count")


```

Good; the "0" dates are gone. The plot has the x axis starting at 1700 -could it be that a few extant buildings from that time are being represented? Yes, that's the case:

```{r}
head(arrange(TAdata[c('ST_NUM', 'ST_NAME', 'ST_NAME_SUF', 'ZIPCODE', 'YR_BUILT_REVISED')], YR_BUILT_REVISED))

```
(it must be mentioned that I took a look at a few of those buildings in Google Street View, and they do not look that old at all... maybe we'll investigate further in the future)



Next, we will create two additional values, that will help us provide a more nuanced analysis of assessed land and building values in Boston, by giving us values normalized by total land area or gross area.

"AV_LAND_PER_SF" will tell us the assessed value per square foot of a parcel lot:

```{r}
TAdata <- transform(TAdata, AV_LAND_PER_SF = ifelse((AV_LAND != 0 & LAND_SF != 0), AV_LAND / LAND_SF, NA))


```

And "AV_BLDG_PER_SF" will give us the assessed value per square foot of a building:

```{r}
TAdata <- transform(TAdata, AV_BLDG_PER_SF = ifelse((AV_BLDG != 0 & GROSS_AREA != 0), AV_BLDG / GROSS_AREA, NA))
```


How can we use these new variables? 

Well, if we compare the total amount of assessed value per neighborhood, we find the downtown area on top, and Hyde Park at the bottom:

```{r}
BRA_PD.value <- ddply(TAdata[(!is.na(TAdata$BRA_PD) & !is.na(TAdata$AV_BLDG)), ], .(BRA_PD), summarise, bldg.value = sum(as.numeric(AV_BLDG)/1000000000), land.value = sum(as.numeric(AV_LAND)/1000000000))

arrange(BRA_PD.value, desc(bldg.value + land.value))
```
(figures expressed in billions of USD)


But if we take a look at the density function for property value by square meter, a different picture emerges:

```{r, warning=FALSE}
ggplot(TAdata[TAdata$BRA_PD != '' & TAdata$AV_BLDG_PER_SF > 1,], aes(AV_BLDG_PER_SF, fill = BRA_PD)) + geom_density(alpha = 0.3)  + xlim(1,1200) + scale_fill_brewer(name="Neighborhood", palette="Paired") + guides(fill=guide_legend(ncol=4)) + theme(legend.title=element_blank(), legend.direction ="vertical",legend.position = "bottom") + ggtitle("Boston: building assessed value by neighborhood") + labs(x="Neighborhood",y="Building value per sqf")

```

Back Bay/Beacon Hill is the area where the building value tends to be higher, as normalized by area; it's hard to tell which neighborhoods come last, since the density curves are superimposing. A box plot may be easier to read:

```{r, warning=FALSE}
ggplot(TAdata[TAdata$BRA_PD != '' & TAdata$AV_BLDG_PER_SF > 1,], aes(x = BRA_PD, y = AV_BLDG_PER_SF, fill = BRA_PD)) + geom_boxplot() + scale_y_log10() + guides(fill=guide_legend(ncol=4)) + theme(legend.title=element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank(), legend.direction ="vertical",legend.position = "bottom") + ggtitle("Boston: building assessed value by neighborhood") + labs(x="Neighborhood",y="Building value per sqf")
```

That helps (slightly). At least we can tell that Mattapan and Roxbury come last.


And what about land value?

```{r, warning=FALSE}

ggplot(TAdata[TAdata$BRA_PD != '' & TAdata$AV_LAND_PER_SF > 1,], aes(x = BRA_PD, y = AV_LAND_PER_SF, fill = BRA_PD)) + geom_boxplot() + scale_y_log10() + guides(fill=guide_legend(ncol=4)) + theme(legend.title=element_blank(), axis.ticks.x = element_blank(), axis.text.x = element_blank(), legend.direction ="vertical",legend.position = "bottom") + ggtitle("Boston: land assessed value by neighborhood") + labs(x="Neighborhood",y="Land value per sqf")
```

Again, Back Bay/Beacon Hill is the area where values are higher than in the rest of the city. Hyde Park (followed by Mattapan) is where the values tend to be lower.


This illustrates how adding additional variables that combine the information of others in useful ways, can help us arrive at a better understanding of of our data.

