---
title: "Mapping ecometrics"
author: "Hector Antonio Vazquez Brust"
date: "November 6, 2015"
output: html_document
---
```{r, echo=FALSE}
options(scipen = 99) # this prevents R from using scientific notation to display large numbers

library(ggplot2)
TAdata <- read.csv('C:/Users/havb/Dropbox/MSUI/Big Data for Cities - PPUA 5262 - 01/R/Boston-Tax-Assesor-Dataset-2015/data/Tax Assessor 2015 - Data.csv', stringsAsFactors = FALSE) #load the Tax Assesor dataset

TAdata <- transform(TAdata, AV_LAND_PER_SF = ifelse((AV_LAND != 0 & LAND_SF != 0), AV_LAND / LAND_SF, NA))

```

##Mapping the relative value of every building in the city

Having decided that we want to measure urban segregation in Boston, we are looking for clues that we can actually observe in our dataset. 

To understand the spatial arrangement of different (and maybe unequal) characteristics of the city's built environment, we will select a variable to put on the map.

Working the Tax Assessor's dataset, building value is one the variables that usually come to mind when choosing what to compare. We have already created a new variable that measures building value normalized by building size ("AV_BLDG_PER_SF"). We will now create and additional variable, to rank from 1 to 10 the comparative value of a particular building compared to the entirety of Boston's building stock. For the statistically minded, what we'll do is calculate the decile every building on our dataset belongs to.


```{r message=FALSE}
library(dplyr)

#Extract a subset from our columns (Only the variables we are interested in)
parcels <- select(TAdata, AV_LAND, AV_BLDG, AV_TOTAL, GROSS_TAX, LAND_SF, YR_BUILT, YR_REMOD, CT_ID_10, X, Y, BRA_PD, AV_LAND_PER_SF, AV_BLDG_PER_SF)

#Add a column with a 1 to 10 ranking (decile) for the parcel's sq foot value
parcels <- mutate(parcels,  BLDG_RANK = ceiling(rank(AV_BLDG_PER_SF,na.last="keep")/length(which(!is.na(AV_BLDG_PER_SF)))*10) )

```

So this is how the new decile/rank variable looks:

```{r}
head(select(parcels, AV_BLDG_PER_SF, BLDG_RANK ))
```


Let's map our new variable over Boston:

```{r message=FALSE, warning=FALSE}
require(ggplot2)
require(ggmap)
Boston <- get_map(location=c(left = -71.193799, bottom = 42.11, right = -70.985746, top = 42.515), color = "bw")
base<- ggmap(Boston)

base + geom_point(data=parcels, aes(x = X, y = Y, color = as.factor(BLDG_RANK), alpha = .5)) + scale_colour_brewer(type = "div", palette = 9) + labs(title = "Boston: building stock value distribution", color='Building value decile \
(higher decile = higher assessed value)', alpha = FALSE) + guides(alpha = FALSE) +
  theme(plot.title = element_text(size = 24, face="bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_text(size = 16, face = "bold"),
        legend.text = element_text(size = 14))

```

This map confirms what the conclusions that we arrived to when we compared neighborhood mean values: Back Bay/Beacon Hill as the most valued, Dorchester, Mattapan, Roxbury as the least. However, mapping the values does provide further insights: it is remarkable how spatially concentrated the high value buildings are, and how close to each other lay the areas with the higher and lesser values.

## Mapping diversity

We will now turn our attention towards diversity. Having given a 1 to 10 ranking to every building in the city, we can use that as a taxonomy for an index of the neighborhood internal variety of building value; which can also be interpreted as an index of variety in property affordability. 

We can borrow a a mathematical tool from the field of biology, where diversity is understood as the variety of "species" in a given environment. Buildings corresponding to different value ranks will be considered as members of different "species" in a given neighborhood environment.

To obtain this index we will use Shannon's algorithm (also known as Shannon-Weaver), of widespread use in ecology studies. The Shannon index H can be calculated as:

$$ - \sum_{p=1}^S p_i \log p_i $$

where $p_i$ is the proportion of species $i$ and $S$ is the total number of species.

We'll make use of R package _vegan_, which includes a function to calculate Shannon's index.  

Since we want to compare diversity at the Census Tract level, which provides better spatial resolution that the neighborhood level, we'll aggregate our data by census tract ID.


```{r message=FALSE}
library(vegan)

#Creating a census tract built environment diversity index
built.environment.diversity <-summarise(group_by(filter(parcels, !is.na(CT_ID_10)), CT_ID_10), 
                                        BUILT_ENV_DIV = diversity(table(BLDG_RANK)),
                                        BUILT_ENV_DIS = var(BLDG_RANK, na.rm = TRUE)/mean(BLDG_RANK, na.rm = TRUE))

summary(built.environment.diversity$BUILT_ENV_DIV)
```

Now we load census data that includes ethnic diversity information -or "ethnic heterogeneity", column EthHet), to merge it with our built environment diversity dataset:

```{r}
demographics <- read.csv('C:/Users/havb/Dropbox/MSUI/Big Data for Cities - PPUA 5262 - 01/R/Census/Tract Census Data.csv', stringsAsFactors = FALSE)

summary(demographics$EthHet)

#Combine Built Environment Diversity and demographics data
bed.demog <- merge(demographics, built.environment.diversity)

```

Now we load a shape file with the geographic extents of Boston's census tracts, turn it into a data frame, and merge it with our data:

```{r message=FALSE}
library(rgdal)

tracts_geo <- readOGR(dsn = 'C:/Users/havb/Dropbox/MSUI/Big Data for Cities - PPUA 5262 - 01/R/shapefiles/Tracts_Boston_2015_BARI', 'Tracts_Boston BARI')

tracts_geo <- fortify(tracts_geo, region = "CT_ID_10")

tracts_geo <- merge(tracts_geo, bed.demog, by.x = 'id', by.y = 'CT_ID_10', all.x = TRUE)

```

Ready to map! First, a map of Boston's built environment diversity, by census tract:

```{r warning=FALSE}
base + geom_polygon(aes(x=long, y=lat, group=group, fill=BUILT_ENV_DIS),
                    data=filter(tracts_geo, (grossrent > 0))) + 
  geom_path(aes(x=long, y=lat, group=group), color = 'gray', data=tracts_geo) +
  scale_fill_gradient(low = "firebrick1", high = "darkseagreen1") +
  labs(title = "Boston: built environment value diversity", fill='Diversity index') +
  theme(plot.title = element_text(size = 24, face="bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_text(size = 16, face = "bold"),
        legend.text = element_text(size = 14))

```


And now a map of ethnic diversity. We'll plot the ethnic diversity index that the US Census Bureau has calculated for us:

```{r warning=FALSE}
base + geom_polygon(aes(x=long, y=lat, group=group, fill=EthHet),
                    data=filter(tracts_geo, (grossrent > 0))) + 
  geom_path(aes(x=long, y=lat, group=group), color = 'gray', data=tracts_geo) +
  scale_fill_gradient(low = "firebrick1", high = "darkseagreen1") +
  labs(title = "Boston: Ethnic diversity", fill='Diversity index') +
  theme(plot.title = element_text(size = 24, face="bold"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        legend.title = element_text(size = 16, face = "bold"),
        legend.text = element_text(size = 14))

```

What we are comparing with the last two maps is, in a sense, ethnic diversity versus diversity in purchasing power. It's hard to visualize a clear pattern; conclusions will have to wait until we can refine our analysis.
